Interpolation of Sparse High-Dimensional Data\thanks {This work was supported by the National Science Foundation Grant CNS-1565314.}

### 2019-09-27 Paper_[2019-02]_NUMA/paper.tex
Updated NUMA paper to fix an error in denominator. Working on dissertation coee still.
Updated with analytic example at end of section 5.

\begin{plainlemma}
  \label{lemma:3}
  Given $f$, $\gamma$, $S$ as in {\it Lemma \ref{lemma:1}}, let $X =
  \{x_0$, $x_1$, $\ldots$, $x_d\}$ $\subset S$ be the vertices of a
  $d$-simplex, and let $\hat f(x) = \langle c, x - x_0 \rangle +
  f(x_0)$, $c \in \mathbb{R}^d$ be the linear function interpolating
  $f$ on $X.$ Let $\sigma_d$ be the smallest singular value of the
  matrix $M = (x_1 - x_0$, $\ldots$, $x_d - x_0)$, and $k =
  \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then
  $$\big\|\nabla f(x_0) - c\big\|_2 \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
\end{plainlemma}

\begin{proofdot}
  Consider $f(x) - \hat f(x)$ along the line segment $z(t) = (1-t)x_0
  + t x_j$, $0 \leq t \leq 1.$ By Rolle's Theorem, for some $0 < \hat
  t < 1$, $\big\langle \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0
  \big\rangle = 0.$ Now
  \begin{align*}
    \big| \big< \nabla f(x_0) - c, x_j - x_0 \big \rangle \big|
        &= \big| \big< \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) + \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0 \big \rangle \big| \\
	&= \big| \big\langle \nabla f(x_0) - \nabla f \big(z(\ \hat t\ )\big), x_j - x_0 \big \rangle \big| \\
	&\leq \big \| \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) \big \|_2 \|x_j - x_0\|_2 \\
	&\leq \gamma \|x_0 - z(\ \hat t\ )\|_2\ \|x_j - x_0\|_2 \\
	&\leq \gamma \|x_j - x_0\|_2^2 \leq \gamma k^2, \\
  \end{align*}
  for all $1 \leq j \leq d.$ Using {\it Lemma \ref{lemma:2}},
  $$\big \| \nabla f(x_i) - c \big \|_2^2 \leq \frac{d}{\sigma_d^2}
  \big( \gamma k^2\big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2
  \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
  \qed
\end{proofdot}

\begin{onetheorem}
  Under the assumptions of {\it Lemma \ref{lemma:1}} and {\it Lemma
    \ref{lemma:3}}, for $z \in S$,
  $$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2.$$
\end{onetheorem}

\begin{proofdot}
  Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d} \gamma k^2 / \sigma_d$ by {\it Lemma \ref{lemma:3}}. Now
  \begin{align*}
    \big|f(z) - \hat f(z)\big|
       &= \big|f(z) - f(x_0) - \langle c, z - x_0 \rangle \big| \\ 
       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) - v, z - x_0 \rangle \big| \\
       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle + \langle v , z - x_0 \rangle \big| \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \big| \langle v , z - x_0 \rangle \big| \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \|v\|_2 \|z - x_0\|_2 \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0), z - x_0 \rangle \big| + \textstyle{\frac{\gamma k^2 \sqrt{d}}{\sigma_d}} \|z - x_0\|_2 \\
       &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d}\frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2,
  \end{align*}
  where the last inequality follows from {\it Lemma \ref{lemma:1}}.
  %
  \qed
\end{proofdot}

In summary, the approximation error of a linear (simplicial)
interpolant tends quadratically towards zero when approaching observed
data only when the diameter of the simplex is also reduced at a
proportional rate.  Only linear convergence to the true function can
be achieved in practice, without the incorporation of additional
observations. Notice that the approximation error is largely
determined by the spacing of observed data. Predictions made by
simplices whose vertices are not well-spaced (i.e., have large
diameter, or are nearly contained in a hyperplane) have higher
error. In light of this error bound, an empirical evaluation of
presented algorithms follows.

>>>>

\begin{plainlemma}
  \label{lemma:3}
  Given $f$, $\gamma$, $S$ as in {\it Lemma \ref{lemma:1}}, let $X =
  \{x_0$, $x_1$, $\ldots$, $x_d\}$ $\subset S$ be the vertices of a
  $d$-simplex, and let $\hat f(x) = \langle c, x - x_0 \rangle +
  f(x_0)$, $c \in \mathbb{R}^d$ be the linear function interpolating
  $f$ on $X.$ Let $\sigma_d$ be the smallest singular value of the
  matrix $M = (x_1 - x_0$, $\ldots$, $x_d - x_0)$, and $k =
  \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then
  $$\big\|\nabla f(x_0) - c\big\|_2 \leq \sqrt{d} \frac{\gamma k^2}{2 \sigma_d}.$$
\end{plainlemma}

\begin{proofdot}
  Consider $f(x) - \hat f(x)$ along the line segment $z(t) = (1-t)x_0
  + t x_j$, $0 \leq t \leq 1.$ By Rolle's Theorem, for some $0 < \hat
  t < 1$, $\big\langle \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0
  \big\rangle = 0.$ Now
  \begin{align*}
    \big| \big< \nabla f(x_0) - c, x_j - x_0 \big \rangle \big|
        &= \big| \big< \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) + \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0 \big \rangle \big| \\
	&= \big| \big\langle \nabla f(x_0) - \nabla f \big(z(\ \hat t\ )\big), x_j - x_0 \big \rangle \big| \\
	&\leq \big \| \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) \big \|_2 \|x_j - x_0\|_2 \\
	&\leq \gamma \|x_0 - z(\ \hat t\ )\|_2\ \|x_j - x_0\|_2 \\
	&\leq \frac{\gamma}{2} \|x_j - x_0\|_2^2 \leq \frac{\gamma k^2}{2}, \\
  \end{align*}
  for all $1 \leq j \leq d.$ Using {\it Lemma \ref{lemma:2}},
  $$\big \| \nabla f(x_i) - c \big \|_2^2 \leq \frac{d}{\sigma_d^2}
  \big( \gamma k^2 / 2 \big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2
  \leq \sqrt{d} \frac{\gamma k^2}{2 \sigma_d}.$$
  \qed
\end{proofdot}

\begin{onetheorem}
  Under the assumptions of {\it Lemma \ref{lemma:1}} and {\it Lemma
    \ref{lemma:3}}, for $z \in S$,
  $$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}{2 \sigma_d} \|z - x_0\|_2.$$
\end{onetheorem}

\begin{proofdot}
  Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d} \gamma k^2 / (2 \sigma_d)$ by {\it Lemma \ref{lemma:3}}. Now
  \begin{align*}
    \big|f(z) - \hat f(z)\big|
       &= \big|f(z) - f(x_0) - \langle c, z - x_0 \rangle \big| \\ 
       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) - v, z - x_0 \rangle \big| \\
       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle + \langle v , z - x_0 \rangle \big| \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \big| \langle v , z - x_0 \rangle \big| \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \|v\|_2 \|z - x_0\|_2 \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0), z - x_0 \rangle \big| + \textstyle{\sqrt{d}\frac{\gamma k^2}{2 \sigma_d}} \|z - x_0\|_2 \\
       &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d}\frac{\gamma k^2}{2 \sigma_d} \|z - x_0\|_2,
  \end{align*}
  where the last inequality follows from {\it Lemma \ref{lemma:1}}.
  %
  \qed
\end{proofdot}

In summary, the approximation error of a linear (simplicial)
interpolant tends quadratically towards zero when approaching observed
data only when the diameter of the simplex is also reduced at a
proportional rate.  Only linear convergence to the true function can
be achieved in practice, without the incorporation of additional
observations. Notice that the approximation error is largely
determined by the spacing of observed data. Predictions made by
simplices whose vertices are not well-spaced (i.e., have large
diameter, or are nearly contained in a hyperplane) have higher
error.

The error bound provided is tight and that is easily observed by
considering the test function $q(x) = \|x\|_2^2,$ $x \in \mathbb{R}^d$
for $d \in \mathcal{N},$ using the simplex $X = \{\mathbf{0}, I\},$
where $I$ is the $d \times d$ identity matrix, and an approximation is
made at $z = \mathbf{1/2}.$ Now this setup yields

\vspace{-4mm}
$$ \gamma = 1, \qquad \sigma_d = 1, \qquad k = 1, \qquad x_0 =
\mathbf{0}, \qquad \hat q(x) = \langle\mathbf{1}, x - \mathbf{0}\rangle +
q(\mathbf{0}). $$
\vspace{-2mm}
\noindent Plugging these values into the Theorem, we get
\vspace{-2mm}

\begin{align*}
  \big|q(z) - \hat q(z)\big| \quad
  =& \quad \big|\|z\|_2^2 - \langle\mathbf{1},z\rangle\big| \\
  =& \quad \bigg| \frac{d}{4} - \frac{d}{2} \bigg| \quad = \quad \frac{d}{4} \\
  \leq& \quad \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}
      {2 \sigma_d} \|z - x_0\|_2 \\
  =& \quad \frac{\|z\|_2^2 + \sqrt{d} \|z\|_2}{2} \\
  =& \quad \bigg(\frac{d}{4} + \frac{d}{4}\bigg)\ /\ 2 \quad = \quad \frac{d}{4}.
\end{align*}

In light of this error bound and analytic test, an empirical
evaluation of presented algorithms follows.
